---
title: "Data Viz Final Project"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(sf)
library(tidyverse)
library(ggiraph)
library(rnaturalearth)
library(rnaturalearthdata)
library(forcats)
library(plotly)
library(scales)
```


```{r, include=FALSE, warning=FALSE}
#read in college scorecard data
scorecard<- read.csv("Most-Recent-Cohorts-All-Data-Elements.csv")

#rename data
scorecard[scorecard == 'NULL'] <- NA
scorecard[scorecard == 'Public'] <- 1
scorecard[scorecard == 'Private nonprofit'] <- 2
scorecard[scorecard == 'Private for-profit'] <- 3
scorecard[scorecard == 'For profit'] <- 3

#split data into male and female only data sets
female_only <- scorecard %>% filter(WOMENONLY == 1)

male_only <- scorecard %>% filter(MENONLY == 1)

male_female_only <- full_join(female_only, male_only)

#male_female_only <- scorecard %>% filter(WOMENONLY == 1 | MENONLY == 1) 

male_female_only <- male_female_only %>% 
 mutate( single_gender_college = ifelse(male_female_only$MENONLY == 1, "Male Only College", "Female Only College"))

make_numeric <- c("COSTT4_A", "SAT_AVG", "ACTMTMID", "FAMINC", "FIRST_GEN", "COSTT4_A", "single_gender_college")

male_female_only <- male_female_only %>% mutate_at(make_numeric, as.numeric)

make_numeric_2 <- c("COSTT4_A", "SAT_AVG", "ACTMTMID", "FAMINC", "FIRST_GEN", "COSTT4_A")
scorecard <- scorecard %>% mutate_at(make_numeric_2, as.numeric)

```
# Investigation of Single Sex Colleges in the U.S. 

## There are 96 active single-sex colleges across the United States, only 35 of which are female-only.  

## Location of Male and Female Only Colleges in US

```{r, tidy=TRUE, echo=FALSE, warning=FALSE}
sf_data <- st_as_sf(male_female_only, coords = c("LONGITUDE", "LATITUDE"), crs = 4326)
library(leaflet)
  leaflet() %>% addTiles() %>% 
  addCircleMarkers(data = sf_data, label = sf_data$INSTNM, fill = sf_data$single_gender_college)
```
#### Most of these colleges are concentrated on the East Coast, and many male only colleges are near majority Jewish communities. Note that all single sex colleges contained in the data are private universities. 

```{r, echo=FALSE, warning=FALSE}
#create map data and plot map of location of male and female only colleges in US
#sf_data <- st_as_sf(male_female_only, coords = c("LONGITUDE", "LATITUDE"), crs = 4326)
#full_scorecard__data <- st_as_sf(scorecard, coords = c("LONGITUDE", "LATITUDE"), crs = 4326)

#scorecard <- scorecard %>% na.omit(LONGITUDE, LATITUDE)

#map <- ggplot() +
  #geom_sf(data = us_states, fill = "white") +
 # geom_sf(data = sf_data, aes(color = single_gender_college)) +
#  geom_point(data = sf_data, aes(text = "INSTNM")) +
  #geom_point(data = full_scorecard__data, aes(color = male_college)) +
 # scale_color_brewer(palette = "Pastel1") +
 # theme_void() +
#  labs(title = "Male and Female Colleges Across the Country")

#ggplotly(map, tooltip = c("text"))
```

## Standardized Test Scores 

### SAT

```{r, tidy=TRUE, echo=FALSE, warning=FALSE}
 SAT_scatter <- ggplot(data = male_female_only,
         aes(x = male_female_only$SAT_AVG, y = male_female_only$FAMINC, color = STABBR, text = text_sat))+
    geom_point(alpha = .5)+
    ggtitle("Family Income and SAT Scores by State")+
    xlab("SAT Score")+
    scale_y_continuous(name = "Family Income",
                       labels = scales::dollar) +
  geom_vline(xintercept = mean(scorecard$SAT_AVG, na.rm = TRUE), color = "violet") +
  geom_vline(xintercept = mean(male_female_only$SAT_AVG, na.rm = TRUE), color = "lightblue") + 
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black")) 

text_sat = paste("State: ", male_female_only$STABBR,"\n Average SAT Score: ", male_female_only$SAT_AVG, sep="")

ggplotly(SAT_scatter, tooltip = "text_sat")
```
#### SAT scores for male and female only colleges lag slightly behind the national average. 

##### Note: The pink line marks national average SAT score, blue line marks average for single-sex colleges. Hover for further stats.

### ACT
```{r, tidy=TRUE, echo=FALSE, warning=FALSE}

male_female_only <- male_female_only %>% filter(ACTMTMID != "NULL")

 ACT_scatter <- ggplot(data = male_female_only,
         aes(x = male_female_only$ACTMTMID, y = male_female_only$FAMINC, color = STABBR, text = text_act))+
    geom_point(alpha = .5)+
    ggtitle("Family Income and ACT Scores by State")+
    xlab("ACT Score")+
    scale_y_continuous(name = "Family Income",
                       labels = scales::dollar) +
    geom_vline(xintercept = mean(scorecard$ACTMTMID, na.rm = TRUE), color = "pink") +
  geom_vline(xintercept = mean(male_female_only$ACTMTMID, na.rm = TRUE), color = "lightblue")+
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black")) 

text_act = paste("State: ", male_female_only$STABBR,"\n Median ACT Score: ", male_female_only$ACTMTMID, sep="")


ggplotly(ACT_scatter, text = "text_act")
```
#### ACT scores for male and female only colleges are slightly above the national average. 

### In both cases, we can see that averages for single sex colleges are overall on par with national averages.

##### Note: The pink line marks national median ACT score, blue line marks average for single-sex colleges. Hover for further stats.


## Share of First Generation Students 

```{r, tidy=TRUE, echo=FALSE, warning=FALSE}
gendered_fg_plot <- 
  male_female_only %>% 
  group_by(STABBR, single_gender_college) %>%
  summarise(mean_fg = mean(FIRST_GEN, na.rm = TRUE))

gendered_fg_plot <-  gendered_fg_plot %>%
  group_by(STABBR) %>%
    select(STABBR, mean_fg, single_gender_college) %>%
    filter(mean_fg != "NaN")


comparison_states <- c("AL", "CA", "DC", "GA", "IN", "MA", "MD", "MN", "MO", "NC", "NE", "NJ", "NY", "OH", "PA", "SC", "VA", "WI")

fg_plot <- scorecard %>% filter(STABBR == "AL" | STABBR == "CA" | STABBR == "DC"| STABBR == "GA"| STABBR == "IN"| STABBR == "MA" | STABBR == "MD" | STABBR == "MN" | STABBR == "MO" | STABBR == "NC" | STABBR == "NE" | STABBR == "NJ"| STABBR == "NY" | STABBR == "OH" | STABBR == "PA" | STABBR == "SC" | STABBR == "VA" | STABBR == "WI") %>% select(STABBR, FIRST_GEN) %>% group_by(STABBR) %>% summarise(mean_first_g = mean(FIRST_GEN, na.rm = TRUE))


first_gen_hist <- ggplot(gendered_fg_plot) +
  geom_histogram(aes(x = STABBR, y = mean_fg, text = text_fg), stat = "identity", fill = "lightblue") +
geom_hline(yintercept = mean(fg_plot$mean_first_g, na.rm = TRUE), color = "pink") +
  geom_hline(yintercept = mean(gendered_fg_plot$mean_fg, na.rm = TRUE), color = "lightblue") +
  labs(x = "State", y = "Mean Share of First Generation Students in Student Body", title = "Mean Share of First Generation Students in Single Sex Colleges") +
theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black")) 

text_fg = paste("State: ", gendered_fg_plot$STABBR, "\nMean Share of First Generation Students: ", gendered_fg_plot$mean_fg, sep="")

ggplotly(first_gen_hist, tooltip = "text_fg")
```
#### Male and female only colleges tend to have more first-generation students compared to the national average. 

##### Note: The pink line marks national average mean share of first-generation students, blue line marks average for single-sex colleges. Hover for further stats.



## Average Cost of Attendance

```{r, echo=FALSE, warning=FALSE}

cost_attendance <- ggplot(data = male_female_only)+
   geom_boxplot(aes(x = STABBR, y = COSTT4_A), fill = "lightblue")+
   ggtitle("Cost of Attendance for Male and Female Only Colleges")+
   xlab("State")+
   scale_y_continuous(name = "Average Cost of Attendance",
                      labels = scales::dollar) +
   geom_hline(yintercept = 47510.47, color = "pink") +
  geom_hline(yintercept = mean(male_female_only$COSTT4_A, na.rm = TRUE), color = "lightblue") +
   labs(x = "State", y = "Average Cost of Attendance", title = "Average Cost of Attendance at Single-Gender Colleges") +
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black")) 


text = paste("State: ", male_female_only$STABBR, "\nFamily Income: ", male_female_only$FAMINC, "\n Cost of Attendance: ", male_female_only$COSTT4_A, sep="") 

ggplotly(cost_attendance, tooltip = "text")
```
#### The cost of male and female only colleges is slightly lower than that of the national average. 

##### Note: The pink line marks national average cost of attendance at private and for profit institutions, blue line marks average for single-sex colleges. Hover for further stats.


###### source: https://collegescorecard.ed.gov/data/ (2022)