---
title: "Data Viz Final Project"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Desktop/Data Visualization/data-viz-final-project")
library(sf)
library(tidyverse)
library(ggiraph)
library(rnaturalearth)
library(rnaturalearthdata)
library(forcats)
library(plotly)
```


```{r, include=FALSE, warning=FALSE}
#read in college scorecard data
scorecard<- read.csv("Most-Recent-Cohorts-All-Data-Elements.csv")

#rename data
scorecard[scorecard == 'NULL'] <- NA
scorecard[scorecard == 'Public'] <- 1
scorecard[scorecard == 'Private nonprofit'] <- 2
scorecard[scorecard == 'Private for-profit'] <- 3
scorecard[scorecard == 'For profit'] <- 3

#split data into male and female only data sets
female_only <- scorecard %>% filter(WOMENONLY == 1)

male_only <- scorecard %>% filter(MENONLY == 1)

male_female_only <- full_join(female_only, male_only)

#male_female_only <- scorecard %>% filter(WOMENONLY == 1 | MENONLY == 1) 

male_female_only <- male_female_only %>% 
 mutate(male_female_only, single_gender_college = ifelse(male_female_only$MENONLY == 1, "Male Only College", "Female Only College"))

make_numeric <- c("COSTT4_A", "SAT_AVG", "FAMINC", "FIRST_GEN", "single_gender_college")

male_female_only <- male_female_only %>% mutate_at(make_numeric, as.numeric)

```
# Investigation of Single Sex Collages in the U.S. 

## Location of Male and Female Only Colleges in US

```{r, tidy=TRUE, echo=FALSE, warning=FALSE}
sf_data <- st_as_sf(male_female_only, coords = c("LONGITUDE", "LATITUDE"), crs = 4326)

library(leaflet)
  leaflet() %>% addTiles() %>% 
  addCircleMarkers(data = sf_data, label = sf_data$INSTNM, fill = sf_data$single_gender_college)
```


```{r, echo=FALSE, warning=FALSE}
#create map data and plot map of location of male and female only colleges in US
#sf_data <- st_as_sf(male_female_only, coords = c("LONGITUDE", "LATITUDE"), crs = 4326)
#full_scorecard__data <- st_as_sf(scorecard, coords = c("LONGITUDE", "LATITUDE"), crs = 4326)

#scorecard <- scorecard %>% na.omit(LONGITUDE, LATITUDE)

#map <- ggplot() +
  #geom_sf(data = us_states, fill = "white") +
 # geom_sf(data = sf_data, aes(color = single_gender_college)) +
#  geom_point(data = sf_data, aes(text = "INSTNM")) +
  #geom_point(data = full_scorecard__data, aes(color = male_college)) +
 # scale_color_brewer(palette = "Pastel1") +
 # theme_void() +
#  labs(title = "Male and Female Colleges Across the Country")

#ggplotly(map, tooltip = c("text"))
```

## Standardized Test Scores 

```{r, tidy=TRUE, echo=FALSE, warning=FALSE}
 SAT_scatter <- ggplot(data = male_female_only,
         aes(x = male_female_only$SAT_AVG, y = male_female_only$FAMINC, color = STABBR, text = text_sat))+
    geom_point(alpha = .5)+
    ggtitle("Family Income and SAT Scores by State")+
    xlab("SAT Score")+
    scale_y_continuous(name = "Family Income",
                       labels = scales::dollar) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  geom_vline(xintercept = mean(scorecard$SAT_AVG, na.rm = TRUE), color = "pink") +
  geom_vline(xintercept = mean(male_female_only$SAT_AVG, na.rm = TRUE), color = "lightblue")

text_sat = paste("State: ", male_female_only$STABBR,"\n Average SAT Score: ", male_female_only$SAT_AVG, sep="")

annotation <- data.frame(
   x = c(1100,1150),
   y = c(75000,75000),
   label = c("National Average", "Single Sex College Average")
)

# Add text
#SAT_scatter <- SAT_scatter + geom_text(data=annotation, aes( x=x, y=y, label=label),                 , 
 #          color="orange", 
  #         size=7 , angle=45, fontface="bold" )

ggplotly(SAT_scatter, tooltip = "text_sat")
```
### note: pink line marks national average SAT Score, blue line marks average for single-sex colleges


```{r, tidy=TRUE, echo=FALSE, warning=FALSE}

male_female_only <- male_female_only %>% filter(ACTMTMID != "NULL")

 ACT_scatter <- ggplot(data = male_female_only,
         aes(x = male_female_only$ACTMTMID, y = male_female_only$FAMINC, color = STABBR, text = text_act))+
    geom_point(alpha = .5)+
    ggtitle("Family Income and ACT Scores by State")+
    xlab("ACT Score")+
    scale_y_continuous(name = "Family Income",
                       labels = scales::dollar) +
    geom_vline(xintercept = mean(scorecard$ACTMTMID, na.rm = TRUE), color = "pink") +
  geom_vline(xintercept = mean(male_female_only$ACTMTMID, na.rm = TRUE), color = "lightblue")+
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black")) 

text_act = paste("State: ", male_female_only$STABBR,"\n Median ACT Score: ", male_female_only$ACTMTMID, sep="")


ggplotly(ACT_scatter, text = "text_act")
```
### note: pink line marks national median ACT Score, blue line marks average for single-sex colleges



## Share of First Generation Students 

```{r, tidy=TRUE, echo=FALSE, warning=FALSE}
gendered_fg_plot <- 
  male_female_only %>% 
  group_by(STABBR) %>%
  summarise(mean_fg = mean(FIRST_GEN, na.rm = TRUE))

gendered_fg_plot <-  gendered_fg_plot %>%
  group_by(STABBR) %>%
    select(STABBR, mean_fg) %>%
    filter(mean_fg != "NaN")


comparison_states <- c("AL", "CA", "DC", "GA", "IN", "MA", "MD", "MN", "MO", "NC", "NE", "NJ", "NY", "OH", "PA", "SC", "VA", "WI")

fg_plot <- scorecard %>% filter(STABBR == "AL" | STABBR == "CA" | STABBR == "DC"| STABBR == "GA"| STABBR == "IN"| STABBR == "MA" | STABBR == "MD" | STABBR == "MN" | STABBR == "MO" | STABBR == "NC" | STABBR == "NE" | STABBR == "NJ"| STABBR == "NY" | STABBR == "OH" | STABBR == "PA" | STABBR == "SC" | STABBR == "VA" | STABBR == "WI") %>% select(STABBR, FIRST_GEN) %>% group_by(STABBR) %>% summarise(mean_first_g = mean(FIRST_GEN, na.rm = TRUE))


first_gen_hist <- ggplot(gendered_fg_plot) +
  geom_histogram(aes(x = STABBR, y = mean_fg, text = text_fg), stat = "identity", fill = "lightblue") +
geom_hline(yintercept = mean(fg_plot$mean_first_g, na.rm = TRUE), color = "pink") +
  geom_hline(yintercept = mean(gendered_fg_plot$mean_fg, na.rm = TRUE), color = "lightblue") +
  labs(x = "State", y = "Mean Share of First Generation Students in Student Body") +
theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black")) 

text_fg = paste("State: ", gendered_fg_plot$STABBR, "\nMean Share of First Generation Students: ", gendered_fg_plot$mean_fg, sep="")

ggplotly(first_gen_hist, tooltip = "text_fg")
```
### note: pink line marks national average mean share of first-generation students, blue line marks average for single-sex colleges



## Average Cost of Attendance

```{r, echo=FALSE, warning=FALSE}

cost_attendance <- ggplot(data = male_female_only)+
   geom_boxplot(aes(x = STABBR, y = COSTT4_A), fill = "lightblue")+
   ggtitle("Cost of Attendance for Male and Female Only Colleges")+
   xlab("State")+
   scale_y_continuous(name = "Average Cost of Attendance",
                      labels = scales::dollar) +
   geom_hline(yintercept = 47510.47, color = "pink") +
   labs(x = "State", y = "Average Cost of Attendance", title = "Average Cost of Attendance at Single-Gender Colleges") +
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"))

text = paste("State: ", male_female_only$STABBR, "\nFamily Income: ", male_female_only$FAMINC, "\n Cost of Attendance: ", male_female_only$COSTT4_A, sep="")


ggplotly(cost_attendance, tooltip = "text")
```
### note: pink line marks national average cost of attendance at private and forprofit institutions


#### source: https://collegescorecard.ed.gov/data/ (2022)